To be FAIR: Theory Specification Needs an Update

### 2025-02-19 manuscript.tex
rephrase paradigm section

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\makeatother
\keywords{fairtheory, meta theory, theory formation, cumulative science, formal models\newline\indent Word count: 8688}
\usepackage{lineno}

\linenumbers
\usepackage{csquotes}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={FAIR theory: Applying Open Science Principles to the Construction and Iterative Improvement of Scientific Theories},
  pdfauthor={Caspar J. Van Lissa1, Aaron Peikert2,3, Andreas M. Brandmaier2,3,4, Felix D. Schönbrodt5, \& Noah N.N. van Dongen6},
  pdflang={en-EN},
  pdfkeywords={fairtheory, meta theory, theory formation, cumulative science, formal models},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{FAIR theory: Applying Open Science Principles to the Construction and Iterative Improvement of Scientific Theories}
\author{Caspar J. Van Lissa\textsuperscript{1}, Aaron Peikert\textsuperscript{2,3}, Andreas M. Brandmaier\textsuperscript{2,3,4}, Felix D. Schönbrodt\textsuperscript{5}, \& Noah N.N. van Dongen\textsuperscript{6}}
\date{}

@@

This is a preprint paper, generated from Git Commit \# c910eaa6.

@@

\abstract{%
Test test.
}

@@

Given these concerns, it is an imbalance that scientific reform initiated by the open science movement has focused primarily on improving deductive methods.
The equally critical inductive processes of theory construction and theory improvement have been largely overlooked\footnote{We use induction to describe inferences from specific observations to general theories; others have coined \emph{abduction} to describe ``inference to the best explanation'' {[}REF Peirce{]}. However, Peirce also acknowledged that abduction ``is not essentially different from induction'' (p.4481). Defining what makes a theory ``best'' requires making auxiliary assumptions. For present purposes, induction and abduction are interchangeable terms.}.
The present paper restores balance by applying, for the first time,
open science principles to psychological theory.
We apply the FAIR principles to scientific theories,
introducing the concept of \emph{FAIR theory} to
facilitate transparent scholarly communication and accelerate cumulative knowledge acquisition.

@@

In a progressive research program (Lakatos, 1971),
this cycle is regularly completed to iteratively advance our understanding of the studied phenomena
.
There are, however, indications that contemporary psychology falls short of this ideal.
Firstly, because hypothesis-testing research is over-represented in the literature:
According to Kühberger, Fritz, and Scherndl (2014), 89.6\% of papers published in psychology report confirmatory hypothesis tests.
In addition, the link between theory and hypothesis is often tenuous or absent (Oberauer \& Lewandowsky, 2019; Scheel, Tiokhin, Isager, \& Lakens, 2021).
Only 15\% of deductive studies referenced any theory, and theory was often not cited in relation to the hypothesis (McPhetres et al., 2021).
The remaining 85\% of deductive studies lacked an explicit connection between theory and hypothesis.
In the best case, such ungrounded hypotheses are rooted in researchers' implicit theories, in which case it is particularly important to make these explicit Norouzi, Kleinberg, Vermunt, \& Van Lissa (2024).
Or, perhaps the hypotheses are not of substantive interest, such as null hypotheses that exist purely for the purpose of being rejected (Van Lissa et al., 2020), and researchers are simply testing them as part of a cultural ritual (Gigerenzer, Krauss, \& Vitouch, 2004).
Testing ad-hoc hypotheses not grounded in theory does not advance our principled understanding of psychological phenomena.
Put differently: collecting significance statements about ad-hoc hypotheses is much like trying to write novels by collecting sentences from randomly generated letter strings (van Rooij \& Baggio, 2021).

@@

As an applied example, consider a comprehensive theory of disease spread and pandemics which covers various psychological factors such as adherence to pandemic mitigation methods (e.g., ), pandemic-related social disruption (e.g., panic buying), or pandemic-related distress and related problems (e.g., anxiety) (Taylor, 2022).
The theory may encompass a particular transmission \emph{model} for disease spread including precise parameters for the process of infection (e.g., social distance, average duration of encounters, ventilation) and incubation times.

@@

Making theories Findable would allow researchers to easily identify relevant theories to inform their hypotheses,
grounding their work in established theoretical foundations.
It further increases the impact and reuse potential of theories across disciplines,
either through direct application (where one discipline stumbles upon a problem that is already well-understood in another discipline),
or through analogical modeling.
In analog modeling, the structure of a theory from one discipline is applied to a phenomenon in another field.
For example, predator-prey models have inspired theories of XXX, and the Eysenck model of atomic magnetism has inspired a network theory of depression.
Findability also enables meta-research on theories,
in the same way libraries and search engines have enabled scholars to study the literature via systematic reviews.
In a similar way, it would become much easier to explicitly compare different theories of a specific phenomenon,
or to study structural properties of theories.

The four Findability criteria are applicable to theory with only minor adjustments, see Table \ref{tab:tabfair}.
First, this requires assigning a globally unique and persistent identifier, or DOI, to each theory (F1).
Of the many services that provide DOIs for scientific information artefacts,
Zenodo and the Open Science Framework are commonly used in psychology.
Second, Findable theory is described with rich metadata (F2).
This includes citation metadata (e.g., referencing a scientific paper that documents the theory, or a psychometric paper that operationalizes specific constructs).
It might further include domain-specific metadata, such as a reference to a taxonomy of psychological constructs (Bosco, Uggerslev, \& Steel, 2017),
ontology (Guyon, Kop, Juhel, \& Falissard, 2018),
or catalog of psychological phenomena {[}REF Noah Denny{]}.
Metadata should also include identifiers for all the versions of the theory it describes (F3);
Zenodo handles this by default by providing an overarching DOI for an information artifact which subsumes the DOIs of that artifact's versions.

Finally, metadata should be registered or indexed in a searchable registry (F4).
Zenodo and GitHub are both searchable.
Standardized metadata further enhance Findability in these repositories.
The DataCite Metadata Schema provides a controlled vocabulary for research output, and the \texttt{resource\_type:\ model} matches the description of FAIR theory (\textbf{datacitemetadataworkinggroupDataCiteMetadataSchema2024?}).
Furthermore, a standard keyword can be used; we suggest using the keyword \texttt{"fairtheory"} for all resources that constitute or reference a FAIR theory.

@@

There have been notable efforts to improve theories' findability through post-hoc curation.
For example, Gray and colleagues introduced a format for representing theories,
and post many examples on their website (Gray, 2017).
Similarly, Borsboom and colleagues seek to establish a database of psychological theories {[}REF BORSBOOM{]}.
Post-hoc curation is a notable effort but does not address the root cause of the lack of Findability, however.
Ideally, Findability would be addressed ante-hoc, through documentation with rich metadata and modular publishing.

@@

Transparent scholarly communication about theory requires that theories are accessible to all researchers and other stakeholders.
If theories are accessible, researchers can reuse and refine them,
thus accelerating cumulative knowledge acquisition.
Making theories accessible also allows stakeholders (e.g., practitioners, policy makers, advocates) to inform themselves of the current scientific understanding of specific phenomena.
While isolated empirical findings can appear fragmented and contradictory (Dumas-Mallet, Smith, Boraud, \& Gonon, 2017),
theories offer a top-down, big picture representation of the phenomena studied in a field.
In other words, theories are an important instrument in science communication.

@@

Successful communication requires shared background knowledge between sender and receiver (Vogt et al., 2024).
The Kuhnian notion of ``normal science'', conducted within the context of a shared paradigm, provides shared background knowledge to facilitate mutual understanding (Kuhn, 2009).
From a pragmatic perspective, these considerations indicate that,
when striving to make theory accessible,
it is important to be as explicit as possible (e.g., about assumptions and ontological definitions),
while acknowledging that accessibility exists on a spectrum,
and that it is impossible to eliminate all ambiguity.
Rather, it may benefit scientific discourse to anticipate misunderstanding,
and use it to drive further explication of theory.
In sum, efforts to communicate theory clearly, with as few dependencies on shared background knowledge as possible, including by formalization, explication of assumptions,
and cross-references to resources that provide relevant context (papers, ontologies, macro-theories, theories of measurement)
will advance its Accessibility.

@@

With regard to the state of interoperability in contemporary psychology,
Kurt Lewin's adage ``there's nothing as practical as a good theory'' (Lewin, 1943) implies that ought to be highly X-interoperable in psychological researchers' day-to-day work.
But, as we argued, this is not the case.
The examples of X-interoperability offered in Table \ref{tab:tabmeehl} illustrate that much can be gained by integrating theory directly into analysis workflows, and by making theory X-interoperable within software used for analysis.
For example, interoperable theory could be used
to select control variables for causal inference (Cinelli, Forney, \& Pearl, 2022),
or to preregister the inferential procedure that would lead to specific modifications of a theory after analyzing empirical data (Peikert et al., 2021),
or to derive machine-readable hypotheses (Lakens \& DeBruine, 2021) which could be automatically evaluated through integration testing (Van Lissa, 2023).
Furthermore, theories can be X-interoperable with each other to enable nesting, or using one theory to clarify elements of another theory.
For example, it should be possible to embed a theory about emotion regulation (e.g., Gross, 2015) within a theory of emotion regulation development (Morris, Silk, Steinberg, Myers, \& Robinson, 2007).

@@

If we take cumulative knowledge acquisition to be a goal of scientific research, then Reusability is the ultimate purpose of making theory FAIR.
Applied to FAIR theory, reusability requires that theory and its associated metadata are richly described with a plurality of accurate and relevant attributes (R1) with a clear and accessible license for reuse (R1.1).
It should further have detailed provenance (R1.2),
which is achieved through version control with Git and archival on Zenodo.
Finally, the (meta)data which meets domain-relevant community standards (R1.3).
The Datacite metadata schema offers an initial template in this regard,
and this paper takes one step towards establishing more fine grained community standards for FAIR theory.

If we consider the current state of Reusability in psychological theory, there appears to be a norm \emph{against} theory reuse:
\emph{``{[}Theories are{]} like toothbrushes --- no self-respecting person wants to use anyone else's''} (Mischel, 2008).
As cumulative knowledge acquisition requires reusable theories that are continuously updated based on insights from new data, such a norm impedes scientific progress (de Groot, 1961).
In FAIR theory workshops, we similarly notice reluctance to reusing and adapting existing theories.
Students ask questions such as ``who owns theory'',
and ``who determines how a theory may be reused or changed''?
These questions imply a norm against modifying theory without consent from the author reminiscent of the aforementioned problem of dependency on the author.

@@

Aside from legal conditions for reuse, there are also social considerations.
For example, while a CC0 license does not mandate attribution,
it is nonetheless essential that scholars comprehensively cite theory and related works to comply with established norms of attribution and comprehensive citation (Aalbersberg et al., 2018).
Another way to guide the social process of (diffuse) collaboration is to include a ``README'' file in the theory repository, which informs users about the ways in which they can reuse and contribute to a FAIR theory.
It is also possible to create or adopt a ``Code of Conduct'' which prescribes

@@

Given that we structured our argument around the importance of FAIR theory for cumulative knowledge production through scientific research around the \emph{empirical cycle},
we decided to use it as an example for this tutorial.
The resulting FAIR theory is available at \url{https://doi.org/10.5281/zenodo.14552329}.
The empirical cycle is described on page 28 of De Groot and Spiekerman (1969).
Note that, while De Groot does not explicitly refer to the empirical cycle as a ``theory'', he derives it from ``a theory of thinking''.
We can thus consider it a meta-theory of theory construction.
The ``empirical cycle'' theory presented by De Groot consists of a series of natural language statements:

@@

If we compare it to the levels of theory formalization (Guest \& Martin, 2021),
it is defined at either the ``theory'' or ``specification'' level.
We can increase the level of formalization, and present an ``implementation'' in the human- and machine-readable DOT language:

@@

This implementation describes the model as a directed graph.
Note that the code has been organized so that the first half describes an ontology of the entities the theory postulates,
and the second half describes their proposed interrelations.
This follows the first two properties of good theory according to Meehl (Paul E. Meehl, 1990).

We can now write this implementation of the empirical cycle to a text file, say \texttt{empirical\_cycle.dot}.

@@

Create a new folder and copy the theory file from the previous step into it.
To help meet the Interoperability and Reusability criteria,
add two more files:
A README.md file with instructions for future users of your theory,
and a LICENSE file with the legal conditions for reuse.
We recommend the \texttt{CC0} license, but other options are available, see \href{https://choosealicense.com/non-software/}{https://choosealicense.com}.

\subsubsection{What's in a README?}\label{whats-in-a-readme}

The readme should contain information to help people get started with using your FAIR theory.
We suggest the following elements:

\begin{itemize}
\tightlist
\item
  Title, prefaced with \texttt{\#\ FAIR\ theory:\ The\ Theory\textquotesingle{}s\ Name}
\item
  Description: A plain-text description of the theory and its scope
\item
  Interoperability: Most README files contain a section labeled ``Getting Started'', ``Instructions'', or ``How to Use''. From a FAIR perspective, such a section might be better labeled ``Interoperability'', or ``How to Use (Interoperability)''. We propose explicitly addressing the theory's X-interoperability, telling users exactly what they can use the theory for, and how. For example, our example is implemented in the DOT language for describing graphs, so we would could provide instructions here on how to plot a DOT graph.
\item
  Contributing: Pertaining to the Reusability criterion, this section should tell users the \emph{social expectations regarding reuse and contributions}.
\item
  License: The legal complement to the preceding section, this section should refer readers to the LICENSE file to learn about the \emph{legal conditions of reuse}.
\item
  Citing this work: Tell users how to cite the theory. Note that this section is redundant with the Zenodo archive, which has a preferred citation field. The disadvantage of redundant information is that you may have to maintain this section of the README going forward. The advantage is that documenting related works in the README makes it more readily accessible to users. We suggest a compromise: to retain this section, but refer the reader to the Zenodo page.
\item
  Related works: This section should refer to the work that the FAIR theory is derived from, or documented in. Again, this is redundant with metadata entered in Zenodo (step 5). We nevertheless recommend using this section to refer to Zenodo, and/or to document one canonical reference for the theory that is unlikely to change going forward. For example, we referenced the original empirical cycle paper here:
\end{itemize}

\begin{verbatim}
This repository contains an implementation of the "empirical cycle",
a model proposed by De Groot and Spiekerman (1969, p. 28). See Zenodo for other related works.

> De Groot, A. D., & Spiekerman, J. A. A. (1969). Methodology:
Foundations of inference and research in the behavioral sciences.
De Gruyter Mouton. https://doi.org/10.1515/9783112313121
\end{verbatim}

@@

The first author was inspired by De Groot too,
but they conceive of the empirical cycle in yet another way.
First, notice that the nodes in De Groot's formulation mostly refer to processes.
This invites the question of what the deliverables are in each phase, or in other words: what actually changes when going through the cycle, except the scholar's mind.
In our implementation below, we account for this difference by having the nodes refer to specific deliverables; the edges now refer to processes.
Second, De Groot's strict distinction between processes of observation, induction, and deduction is not widely supported by philosophy of science.
For example, many have argued that observation is value-laden, and as such, involves induction.
The derivation of hypotheses from theory is also not purely deductive,
as auxiliary assumptions are often made (which are, again, an inductive process).
Furthermore, if the testing procedure is not explicitly defined before seeing the data, it incurs some inductive bias as well {[}REF Peikert{]}.
With these alterations, we implement the empirical cycle as follows:

@@

Some have argued that \emph{causal explanations} are a property of good theory {[}REF Meehl, etc?{]}.
According to Pearl and colleagues,
explicit assumptions about the direction of causality allow one to perform causal inference even on cross-sectional data.
Any formal theory that is explicit about direction of causality could thus be used to guide causal inference,
and could even be integrated into the analysis environment.

@@

The replication crisis has brought the inadequacies of contemporary theoretical practices in the social sciences into focus.
Psychological theories often fall short of all FAIR principles: they are hard to find and access, have limited interoperability, and are rarely reused.
These limitations impede cumulative knowledge production in our field,
leading to an accumulation of ``one-shot'' empirical findings, without commensurate advancement in our principled understanding of psychological phenomena.
We argued that applying the FAIR principles to theory offers a structured solution to these shortcomings.
We demonstrated how to create, version-control, and archive theories as digital information artifacts.
We introduced the \texttt{theorytools} R-package to partly automate these processes, reducing barrier of entry for researchers,
and creating a FAIR resource for theory construction tools and documentation that can be continuously updated as best practices develop further.

Making theory FAIR allows researchers to more easily find a relevant framework;
access and understand it; interact with it in a very practical manner, for example, by deriving predictions from it, or using it to select control variables; and reuse it, contributing changes to existing theories or splitting of in a new direction.
Whereas the idea of theory can be quite nebulous to empirical social scientists,
FAIR theory makes theoretical work practical and tangible, incorporating theory into scholars' workflows.
Having a concrete object to iterate upon facilitates the systematic improvement and iterative refinement of psychological theories, thus substantially increasing the efficiency of research.
While FAIR theory does not directly reduce ambiguity,
it provides a framework within which scholars can iteratively increase precision and formalization.
FAIR principles also facilitates new ways of collaboration,
leveraging tools like Git for version control and Zenodo for archiving to document provenance and facilitate contributions from diverse researchers.

@@

Interdisciplinarity benefits from FAIR theory's accessibility across different fields; thus, theoretical frameworks can be reused, adapted, or used for analogical modeling {[}REF Oisin paper{]}.
Meta-research benefit from the fact that FAIR theory enables studying the structure, content, and development of theories over time.
In terms of team science, FAIR theory facilitates collaboration by ensuring that all contributors have access to the same information and
clarifying any remaining areas of contention or misunderstanding.
Version control provides a framework to resolve parallel developments from multiple collaborators in a non-destructive manner.
This facilitates collaboration across geographical boundaries,
and adversarial collaboration, where others strive to falsify a theory or identify its inconsistencies, and democratizes collaboration with as-of-yet unknown collaborators via platforms like GitHub, where researchers outside one's network can identify issues or suggest improvements to theories.

>>>>

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\makeatother
\keywords{fairtheory, meta theory, theory formation, cumulative science, formal models\newline\indent Word count: 8186}
\usepackage{lineno}

\linenumbers
\usepackage{csquotes}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={To be FAIR: Theory Specification Needs an Update},
  pdfauthor={Caspar J. Van Lissa1, Aaron Peikert2,3, Andreas M. Brandmaier2,3,4, Felix D. Schönbrodt5, \& Noah N.N. van Dongen6},
  pdflang={en-EN},
  pdfkeywords={fairtheory, meta theory, theory formation, cumulative science, formal models},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{To be FAIR: Theory Specification Needs an Update}
\author{Caspar J. Van Lissa\textsuperscript{1}, Aaron Peikert\textsuperscript{2,3}, Andreas M. Brandmaier\textsuperscript{2,3,4}, Felix D. Schönbrodt\textsuperscript{5}, \& Noah N.N. van Dongen\textsuperscript{6}}
\date{}

@@

This is a preprint paper, generated from Git Commit \# c8b13da7.

@@

\abstract{%
Many meta‐scientific initiatives have focused on enhancing the rigor of theory testing, yet the systematic specification of theories remains underdeveloped. We present FAIR theory, a new approach that transforms scientific theories into digital, evolving artifacts---developed collaboratively---to bolster the robustness and interconnectedness of scientific knowledge. By representing theories as formal, machine‐readable objects (e.g., mathematical models, computer code, nomological networks, or explicit operational definitions) and tracking their evolution through version control and semantic versioning, FAIR theory provides a structured, cumulative framework for theory development. We propose that each theory is assigned a universally unique identifier and enriched with detailed metadata, ensuring it is Findable, Accessible, Interoperable, and Reusable. This digital infrastructure addresses a critical gap in scholarly practice by introducing the collaborative tools and standards for theory specification that have already revolutionized methodology, data sharing, and publishing. Ultimately, FAIR theory aims to transform how scientific ideas are communicated, evaluated, and iteratively refined, thereby accelerating cumulative knowledge acquisition across disciplines.
}

@@

Given these concerns, it is an imbalance that scientific reform initiated by the open science movement has focused primarily on improving deductive methods.
The equally critical inductive processes of theory construction and theory improvement have been largely overlooked\footnote{We use induction to describe inferences from specific observations to general theories; others have coined \emph{abduction} to describe ``inference to the best explanation'' (\textbf{peirceCollectedPapersCharles1960?}). However, Peirce also acknowledged that abduction ``is not essentially different from induction'' (p.4881). Defining what makes a theory ``best'' requires making auxiliary assumptions. For present purposes, induction and abduction are interchangeable terms.}.
The present paper restores balance by applying, for the first time,
open science principles to psychological theory.
We apply the FAIR principles to scientific theories,
introducing the concept of \emph{FAIR theory} to
facilitate transparent scholarly communication and accelerate cumulative knowledge acquisition.

@@

In a progressive research program (Lakatos, 1971),
this cycle is regularly completed to iteratively advance our understanding of the studied phenomena
.
There are, however, indications that contemporary psychology falls short of this ideal.
Firstly, because hypothesis-testing research is over-represented in the literature:
According to Kühberger, Fritz, and Scherndl (2014), 89.6\% of papers published in psychology report confirmatory hypothesis tests.
Closer examination of deductive research reveals, however, that the link between theory and hypothesis is often tenuous or absent (Oberauer \& Lewandowsky, 2019; Scheel, Tiokhin, Isager, \& Lakens, 2021).
Only 15\% of deductive studies referenced any theory, and theory was often not cited in relation to the hypothesis (McPhetres et al., 2021).
The remaining 85\% of deductive studies lacked an explicit connection between theory and hypothesis.
In the best case, such ungrounded hypotheses are rooted in researchers' implicit theories, in which case it is particularly important to make these explicit Norouzi, Kleinberg, Vermunt, \& Van Lissa (2024).
Or, perhaps the hypotheses are not of substantive interest, such as null hypotheses that exist purely for the purpose of being rejected (Van Lissa et al., 2020), and researchers are simply testing them as part of a cultural ritual (Gigerenzer, Krauss, \& Vitouch, 2004).
Testing ad-hoc hypotheses not grounded in theory does not advance our principled understanding of psychological phenomena.
Put differently: collecting significance statements about ad-hoc hypotheses is much like trying to write novels by collecting sentences from randomly generated letter strings (van Rooij \& Baggio, 2021).

@@

As an applied example, consider a comprehensive theory of disease spread and pandemics which covers various psychological factors
such as adherence to infection prevention protocols (e.g., social distancing),
pandemic-related behavior (e.g., panic buying),
and pandemic-related distress
(Taylor, 2022).
The theory may encompass a particular transmission \emph{model} for disease spread including precise parameters for the process of infection (e.g., social distance, average duration of encounters, ventilation) and incubation times.

@@

Making theories Findable would allow researchers to easily identify relevant theories
and ground their hypotheses in established theoretical foundations.
It further increases the impact and reuse potential of theories across disciplines,
either through direct application (where one discipline stumbles upon a problem that is already well-understood in another discipline),
or through analogical modeling.
In analogical modeling, the structure of a theory from one discipline is applied to a phenomenon in another field.
For example, predator-prey models have inspired theories of intelligence (\textbf{vandermaasDynamicalModelGeneral2006?}), and the Eysenck model of atomic magnetism has inspired a network theory of depression (\textbf{cramerMajorDepressionComplex2016?}).
Findability also enables meta-research on theories,
in the same way libraries and search engines have enabled scholars to study the literature via systematic reviews.
In a similar way, it would become much easier to explicitly compare different theories of a specific phenomenon,
or to study structural properties of theories.

The four Findability criteria are applicable to theory with only minor adjustments, see Table \ref{tab:tabfair}.
First, this requires assigning a globally unique and persistent identifier, such as a DOI, to each theory (F1).
Of the many services that provide DOIs for scientific information artefacts,
Zenodo and the Open Science Framework are commonly used in psychology.
Second, Findable theory is described with rich metadata (F2).
This includes citation metadata (e.g., referencing a scientific paper that documents the theory, or a psychometric paper that operationalizes specific constructs).
It might further include domain-specific metadata, such as a reference to a taxonomy of psychological constructs (Bosco, Uggerslev, \& Steel, 2017),
ontology (Guyon, Kop, Juhel, \& Falissard, 2018),
or catalog of psychological phenomena {[}REF Noah Denny{]}.
Metadata should also include identifiers for all the versions of the theory it describes (F3);
Zenodo handles this by default by providing an overarching DOI for an information artifact which subsumes the DOIs of that artifact's versions.

Finally, metadata should be registered or indexed in a searchable registry (F4).
Zenodo and GitHub are both searchable.
Standardized metadata further improve Findability in these repositories.
The DataCite Metadata Schema provides a controlled vocabulary for research output, and the \texttt{resource\_type:\ model} matches the description of FAIR theory (\textbf{datacitemetadataworkinggroupDataCiteMetadataSchema2024?}).
Furthermore, a standard keyword can be used; we suggest using the keyword \texttt{"fairtheory"} for all resources that constitute or reference a FAIR theory.

@@

There have been notable efforts to improve theories' findability through post-hoc curation.
For example, Gray and colleagues introduced a format for representing theories,
and post many examples on their website (Gray, 2017).
Similarly, Van Dongen and colleagues are working on a database of models and formalized theories.
Post-hoc curation is a notable effort but does not address the root cause of the lack of Findability, however.
Ideally, Findability would be addressed ante-hoc, through documentation with rich metadata and modular publishing.

@@

Transparent scholarly communication about theory requires that theories are accessible to all researchers and other stakeholders.
If theories are not accessible, researchers cannot reuse and refine them.
Thus, accessibility can accelerate cumulative knowledge acquisition.
Making theories accessible also allows stakeholders (e.g., practitioners, policymakers, advocates) to inform themselves of the current scientific understanding of specific phenomena.
While isolated empirical findings can appear fragmented and contradictory (Dumas-Mallet, Smith, Boraud, \& Gonon, 2017),
theories offer a top-down, big-picture representation of the phenomena studied in a field.
In other words, theories are an important instrument in science communication.

@@

Successful communication requires shared background knowledge between sender and receiver (Vogt et al., 2024).
The Kuhnian notion of ``normal science'', conducted within the context of a shared paradigm, provides shared background knowledge to facilitate mutual understanding (Kuhn, 2009).
From a pragmatic perspective, these considerations indicate that,
when striving to make theory accessible,
it is important to be as explicit as possible (e.g., about assumptions and ontological definitions),
while acknowledging that accessibility exists on a spectrum,
and that it is impossible to eliminate all ambiguity.
Rather, it may benefit scientific discourse to anticipate misunderstanding,
and use it to drive further explication of theory.
In sum, efforts to communicate theory clearly, with as few dependencies on shared background knowledge as possible, including by formalization, explication of assumptions,
and cross-references to resources that provide relevant context (papers, ontologies, macro-theories, theories of measurement)
will advance its Accessibility (\textbf{langeChecklist2025?}).

@@

With regard to the state of interoperability in contemporary psychology,
Kurt Lewin's adage ``there's nothing as practical as a good theory'' (Lewin, 1943) implies that theories ought to be highly X-interoperable in psychological researchers' day-to-day work.
But, as we argued, this is not the case.
The examples of X-interoperability offered in Table \ref{tab:tabmeehl} illustrate that much can be gained by integrating theory directly into analysis workflows, and by making theory X-interoperable within software used for analysis.
For example, interoperable theory could be used
to select control variables for causal inference (Cinelli, Forney, \& Pearl, 2022),
or to preregister the inferential procedure that would lead to specific modifications of a theory after analyzing empirical data (Peikert et al., 2021),
or to derive machine-readable hypotheses (Lakens \& DeBruine, 2021) which could be automatically evaluated through integration testing (Van Lissa, 2023).
Furthermore, theories can be X-interoperable with each other to enable nesting, or using one theory to clarify elements of another theory.
For example, it should be possible to embed a theory about emotion regulation (e.g., Gross, 2015) within a theory of emotion regulation development (Morris, Silk, Steinberg, Myers, \& Robinson, 2007).

@@

If we take cumulative knowledge acquisition to be a goal of scientific research, then Reusability is the ultimate purpose of making theory FAIR.
Applied to FAIR theory, reusability requires that each theory and its associated metadata are richly described with a plurality of accurate and relevant attributes (R1) with a clear and accessible license for reuse (R1.1).
It should further have detailed provenance (R1.2),
which is achieved through version control with Git and archival on Zenodo.
Finally, the (meta)data which meets domain-relevant community standards (R1.3).
The Datacite metadata schema offers an initial template in this regard,
and this paper takes one step towards establishing more fine-grained community standards for FAIR theory.

If we consider the current state of Reusability in psychological theory, there appears to be a norm \emph{against} theory reuse:
\emph{``{[}Theories are{]} like toothbrushes --- no self-respecting person wants to use anyone else's''} (Mischel, 2008).
As cumulative knowledge acquisition requires reusable theories that are continuously updated based on insights from new data, such a norm impedes scientific progress (de Groot, 1961).
In FAIR theory workshops, we similarly notice a reluctance to reuse and adapt existing theories.
Students ask questions such as ``Who owns a theory?'',
and ``Who determines how a theory may be reused or changed?''.
These questions imply a norm against modifying theory without consent from the author. \textless!-- NvD: I suggest removing the rest of the sentence (as a result of an earlier suggestion). -- \textgreater{} reminiscent of the aforementioned problem of dependency on the author.

@@

Aside from legal conditions for reuse, there are also social considerations.
For example, while a CC0 license does not legally mandate attribution,
the norms of good scientific practice require that scholars comprehensively cite theory and related works (Aalbersberg et al., 2018).
Another way to guide the social process of (diffuse) collaboration is to include a ``README'' file in the theory repository, which informs users about the ways in which they can reuse and contribute to a FAIR theory.
It is also possible to create or adopt a ``Code of Conduct'' which prescribes behavioral norms for contributors and users of a theory.

@@

We will use the \emph{empirical cycle} as a running example for this tutorial.
The empirical cycle, described on page 28 of De Groot and Spiekerman (1969),
is a kind of meta-theory of theory construction.
The resulting FAIR implementation of this theory is available at \url{https://doi.org/10.5281/zenodo.14552329}.
The original theory consists of a series of natural language statements:

@@

If we compare it to the levels of theory formalization (Guest \& Martin, 2021),
it is defined at either the ``theory'' or ``specification'' level.
We can increase the level of formalization, and present an ``implementation'' in the human- and machine-readable DOT language (and thereby fulfill criterion I1 of Table \ref{tab:tabfair}):

@@

This implementation describes the model as a directed graph.
Note that the code has been organized so that the first half describes an ontology of the entities the theory postulates,
and the second half describes their proposed interrelations.
This follows the first two properties of good theory according to Meehl (Paul E. Meehl, 1990).
We can now write this implementation of the empirical cycle to a text file, say \texttt{empirical\_cycle.dot}.

@@

Create a new folder and copy the theory file from the previous step into it.
To help meet the Interoperability and Reusability criteria,
add two more files:
A README.md file with instructions for future users of your theory,
and a LICENSE file with the legal conditions for reuse.
For guidance on writing the README file, see \href{https://cjvanlissa.github.io/theorytools/articles/readme.html}{this vignette}
We recommend the \texttt{CC0} license, but other options are available, see \href{https://choosealicense.com/non-software/}{https://choosealicense.com}.

@@

The first author was inspired by De Groot as well,
but they specify the empirical cycle differently.
First, notice that the nodes in De Groot's formulation mostly refer to processes.
This invites the question of what the deliverables are in each phase, or in other words: what actually changes when going through the cycle, except the scholar's mind.
In our implementation below, we account for this difference by having the nodes refer to specific deliverables; the edges now refer to processes.
Second, De Groot's distinction between phases of observation, induction, and deduction is not fully congruent with philosophy of science.
For example, many have argued that observation is theory-laden, and as such, involves induction (\textbf{brewerTheoryLadennessObservationTheoryLadenness2001?}).
Deriving hypotheses from theory is also not purely deductive,
as auxiliary assumptions are often made (which, again, involves induction).
Furthermore, if the testing procedure is not explicitly defined before seeing the data, it incurs some inductive bias as well (\textbf{peikertTransparencyOpenScience2023?}).
With these alterations, we implement the empirical cycle as follows:

@@

Some have argued that \emph{causal explanations} are a property of good theory (Paul E. Meehl, 1990).
According to Pearl and colleagues,
explicit assumptions about the direction of causality allow one to perform causal inference even on cross-sectional data (\textbf{pearlCausalDiagramsEmpirical1995?}).
Any formal theory that is explicit about the direction of causality could thus be used to guide causal inference,
and could even be integrated into the analysis environment.

@@

The replication crisis has brought the inadequacies of contemporary theoretical practices in the social sciences into focus.
Psychological theories often fall short of all FAIR principles: they are hard to find and access, have limited interoperability, and are rarely reused.
These limitations impede cumulative knowledge production in our field,
leading to an accumulation of ``one-shot'' empirical findings, without commensurate advancement in our principled understanding of psychological phenomena.
We argued that applying the FAIR principles to theory offers a structured solution to these shortcomings.
We demonstrated how to create, version-control, and archive theories as digital information artifacts.
We introduced the \texttt{theorytools} R-package to partly automate these processes, reducing the barrier of entry for researchers,
and creating a FAIR resource for theory construction tools and documentation that can be continuously updated as best practices develop further.

Making theory FAIR allows researchers to more easily find a relevant framework;
access and understand it; interact with it in a very practical manner, for example, by deriving predictions from it, or using it to select control variables; and reuse it, contributing changes to existing theories or splitting of in a new direction.
Whereas the idea of theory can be quite nebulous to empirical social scientists,
FAIR theory makes theoretical work practical and tangible, incorporating theory into scholars' workflows.
Having a concrete object to iterate upon facilitates the systematic improvement and iterative refinement of psychological theories, thus substantially increasing the efficiency of research.
While FAIR theory does not directly reduce ambiguity,
it provides a framework within which scholars can iteratively increase precision and formalization.
The FAIR principles also facilitate new ways of collaboration,
leveraging tools like Git for version control and Zenodo for archiving to document provenance and facilitate contributions from diverse researchers.

@@

Interdisciplinarity benefits from FAIR theory's accessibility across different fields; thus, theoretical frameworks can be reused, adapted, or used for analogical modeling {[}REF Oisin paper{]}.
Meta-research benefits from the fact that FAIR theory enables studying the structure, content, and development of theories over time.
In terms of team science, FAIR theory facilitates collaboration by ensuring that all contributors have access to the same information and
clarifying any remaining areas of contention or misunderstanding.
Version control provides a framework to resolve parallel developments from multiple collaborators in a non-destructive manner.
This facilitates collaboration across geographical boundaries,
and adversarial collaboration, where others strive to falsify a theory or identify its inconsistencies, and democratizes collaboration with as-of-yet unknown collaborators via platforms like GitHub, where researchers outside one's network can identify issues or suggest improvements to theories.
